sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-11 doc/paper.Rmd
Editing

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

@@

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and 'group', respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

```{r diagram, fig.cap="(ref:diagram-fig)"}
dir.create("figs", showWarnings = FALSE)
f <- here::here("figs", "diagram.pdf")
if (!file.exists(f)) {
  download.file("https://www.dropbox.com/s/j8km3hv8xz9b63k/diagram.pdf?dl=1", 
    destfile = f)
}
knitr::include_graphics(f)
```

@@

<!-- or areal/grid data with centroids instead of CAR/SAR models -->
<!-- add CAR/SAR to table -->
sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates. 
The data can be collected across discrete points in time.
Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
sdmTMB can also fit models with areal data if each polygon has an associated centroid.
<!-- (observations are treated as replicates). -->
A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range---the distance at which points are effectively independent).

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram)).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).

<!-- highlight where each one excels ... -->
<!-- no CV built into VAST or priors -->
<!-- discussion integrates across speed, familiarity and table 1 -->
<!-- results paragraph on Table 1 -->
<!-- familiarity is an important word here -->
<!-- add section on functionality comparison -->

Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
the general form of a GLMM is
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
Although in practice it is rare to include all of these components, the full sdmTMB model is
$$
g(\mu_{\bm{s},t}) =
\bm{X}_{1,\bm{s},t} \bm{\beta} +
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t +
\bm{X}_{3,t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t},
$$
where

* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.

>>>>

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

@@

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

```{r diagram, fig.cap="(ref:diagram-fig)"}
dir.create("figs", showWarnings = FALSE)
f <- here::here("figs", "diagram-2021-12-10.pdf")
if (!file.exists(f)) {
  download.file("https://www.dropbox.com/s/lvoem3ahwkftzjy/diagram-2021-12-10.pdf?dl=1", 
    destfile = f)
}
knitr::include_graphics(f)
```

@@

sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates. 
The data can be collected across discrete points in time.
Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
sdmTMB can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram)).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
See Appendix TODO for a full description of the statistical model.

<!-- Using the notation of a bold capital letter to represent a data matrix or vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time, -->
<!-- the general form of a GLMM is -->
<!-- $g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$, -->
<!-- with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$. -->
<!-- We can partition the random effects $\bm{Z \psi}$ into subcomponents, each of which can optionally be included in the model (Appendix TODO). -->
<!-- Although, in practice it is rare to include all of these components, the full sdmTMB model is -->
<!-- $$ -->
<!-- g(\mu_{\bm{s},t}) = -->
<!-- \bm{X}_{1,\bm{s},t} \bm{\beta} + -->
<!-- \bm{\alpha}_g + -->
<!-- \bm{X}_{2,\bm{s},t} \gamma_t + -->
<!-- \bm{X}_{3,t} \zeta_{\bm{s}} + -->
<!-- \omega_{\bm{s}} + -->
<!-- \epsilon_{\bm{s},t}, -->
<!-- $$ -->
<!-- where -->

<!-- * $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices; -->
<!-- * $\bm{\beta}$ represents a vector of fixed-effect coefficients; -->
<!-- * $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; -->
<!-- * $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$; -->
<!-- * $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$; -->
<!-- * $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and -->
<!-- * $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field). For the IID case, $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$ (Appendix TODO). -->

@@

By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix TODO).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix TODO).

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
Derived parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.
